Enabling Ambient Backscatter Using Low-cost Software-defined-radio

### 2017-04-06 paper.tex
Some corrections in the design section

\subsection{Transmitter}
\begin{figure}[h]
\centering
\includegraphics[width=0.8\columnwidth]{./fig/s11v2}
\caption{Input reflection coefficient \ensuremath{\Gamma_{in}} also known as \ensuremath{S_{\text{11}}} of the selfmade backsacttering ground plane antenna in dB.}
\label{fig:s11}
\end{figure}
A backscattering antenna is the most important part of backscattering sender. Our antenna is optimized for usage with the television signal, which has its center frequency at 626 MHz. It is a ground plane antenna, which has been built up on a big PCB. This rectangular PCB, with dimensions of 200 mm x 300 mm, is covered with a copper layer on one side and left blank on the other. A hole in the midle of the PCB serves for connecting a thicker wire to a 50 \ensuremath{\Omega} microstrip transmission line. The wire is the vertical antenna element has a length of 350 mm, which is roughly 0.73 \ensuremath{\lambda}. It is soldered to the transmission line. The transmission line has been calculated for 626 MHz using FR4 substrate with a thickness of 1.6 mm. According to microstrip line theory the length of the line changes the phase and the width of the line changes the impedance (cf. chapter 3 in \cite{pozar_microwave_2011}). The design equations given in \cite{pozar_microwave_2011} are implemented in various programs e.g. KiCAD, which has been used for solving the design equations. The given parameters lead to a microstrip line width of 3 mm. An open-stub, attached directly at the puncture where the vertical antenna element is attached to the transmission line, is used for fine tuning of the input impedance. The stub length has been adjusted manually (shortened) while measuring the reflection coefficient with a vector network analyzer (VNA). Open-stub matching is discussed in chapter 5 in \cite{pozar_microwave_2011}. Furthermore the resonance frequency has been tuned by cutting the vertical antenna element piece by piece while measuring with the VNA. At some point the \ensuremath{S_{\text{11}}} had the minimum quite close to where we wanted to have it (cf. \ref{fig:s11}). In the plot one can also see, that the antenna is quite narrowband. For backscattering this is not necessarily a disadvantage, since everything, which is within the operating frequency of the antenna is backscattered. So with a narrowband antenna the effect, that signals are backscattered, which are not targeted, is reduced. This statement has to be qualified as the antenna also has multiple resonance frequencies in multiple bands. With our approach we were able to achieve a input resistance at 634 MHz of \ensuremath{Z_{in}=(52.2 + j 4.3) \Omega}. At 626 MHz \ensuremath{S_{\text{11}}} is still approximately -17 dB.                   
Besides the antenna the transmitter consists out of a \textit{MSP430}, which acts as subscriber, that sends data to the receiver. This transmitter is connected with one I/O pin and ground to a rf switch. The switch can change between matched (50 \ensuremath{\Omega}) and open. The software now steers this switch with a two 2 MHz rectangular timer signal for a 1 and is just leavs it open for a 0. Therefore the result is a shift of the television signal by 2 MHz for a 1 and no shift for a 0. Hence a classical binary amplitude shift keying is the implemented modulation technique.

@@

\caption{Receiver architecture from a system point of view. Sinal flow is from the left to the right.}

@@

The receiver was the most work intensive part of the design. C++ has been used to implement the receiver. We provide the code publicly available under \url{https://github.com/s3xm3x/backscatterBASKReceiver}. In figure \ref{fig:receiver_arch} one can see the architecture of the receiver from an abstarct point of view. A highly sophisticated bus system has been developed to exchange data between the different components of the system. The signal flow is from the left to the right. \textit{Librtlsdr} (cf. \cite{steve-m_librtlsdr}), which is based on \textit{libusb}, is used to transfer the data from the TV stick into our program. As described under \ref{sub:rtl2832}, the RTL2832 mixes down the high frequency to a intermediate frequency (both values are in SW adjustable). Subsequently the data is mixed with sine and cosine, low-pass filtered and finally transfered in the digital world. This data is available as two 8 Bit values. It is a well-known fact, that this technique, which is commonly known as I/Q demodulation, results in a real and an imagenery part. With this two values phase and absolute value of the data can be determined for every sample. So every sample consists out of a real and and imaginary part. And can be seen as:

@@

	I+j\,Q = abs(I,Q) \cdot e^{ang(I,Q)} 

@@

So the first block, which is entitled with RTL2832U, provides the interface to the TV dongle. This block is represented as one object. It can set the frequency \ensuremath{f_{tuned}}, where the receiver is listening. Also it can set the sampling frequency \ensuremath{f_{samp}} and the analog gain of the tuner. The raw samples are pushed to the bus. 
Demodulator is the block called, which is responsible to convert the sampled signals into something, which looks like ones and zeroes. Thereforre the real and imgainary values first have to be deinterleaved out of the sample message. Now they are represented as integer values between 0 and 255 (8 bit). So one converts them to floats, whereas 127.5 equals 0. After that the values have to be downsampled. Usually we were sampling with 250 kHz. This is still much more than we need assuming an additive white gausian noise (AWGN) channel with a \ensuremath{S/N = 10\,dB} (cf. equation \ref{eq:awgn}). So we decided to reduce the sampling frequency to 25 kHz.

@@

A FIR filter naivly implemented with floats has been used to achieve this. 10 kHz is the cut-off frequency of the first filter. All filters have been constructed with the internal filter design tool of GNU radio. After downsampling another filter is used to suppress noise. This filter has a cut-off frequency of 3 kHz. The last demodulation step is to take the signal, rectify it (cf. equation \ref{eq:rectify}) and decide with a software-defined Schmitt trigger whether a sample is a 1 or a 0. 

@@

These samples are broadcasted on the bus. A registered listener receives the processed samples. This registered listener is the decoder, which is represented as an object as well. After the decoder received the samples it decides when a frame is sent or when the channel is idle. On the basis of a defined threshold the decoder determines wether a certain sample pattern of ones and zeros is a 1 bit or a 0 bit. The baudrate and the sampling rate lead to a certain length of one bit. For example a sample rate \ensuremath{f_{sampling}} of 25 kHz combinded with a baudrate of 1 kbit/s leads to 25 samples/bit. The decoder also includes a function to correlate the received bitstream with an expected pattern. Hence it is able to determine the bit ro error ratio (BER). 
Furthermore the code we have written so far subsumes a simulator to simulate ideal data coming out of the demodulator to test the algorithms behind the decoder. Also a simulator for the RTL2832U has been written to replay recorded samples, which facialiates debugging as well.    

>>>>

\subsection{Transmitter}
\begin{figure}[h]
\centering
\includegraphics[width=0.8\columnwidth]{./fig/s11v2}
\caption{Input reflection coefficient \ensuremath{\Gamma_{in}} also known as \ensuremath{S_{\text{11}}} of the self-made backscattering ground-plane antenna.}
\label{fig:s11}
\end{figure}
A backscattering antenna is the most important part of backscatter transmitter. Our antenna is optimized for usage with the television signal, which has its center frequency at 626 MHz. It is a ground plane antenna, which has been built up on a large unetched PCB plate. This rectangular PCB, with dimensions of 200 mm x 300 mm, is covered with a copper layer on one side and left blank on the other. A hole in the middle of the PCB serves for connecting a wire to a 50 \ensuremath{\Omega} micro strip transmission line. The wire, which acts as the vertical antenna element and has a length of 350 mm (roughly 0.73 \ensuremath{\lambda}), is soldered to the transmission line. The transmission line has been calculated for 626 MHz using an FR4 substrate with a thickness of 1.6 mm. According to micro strip line theory the length of the line changes the phase and the width of the line changes the impedance (cf. Chapter 3 in \cite{pozar_microwave_2011}). The design equations given in \cite{pozar_microwave_2011} are implemented in various programs e.g. KiCAD, which has been used for solving the design equations. The given parameters lead to a micro strip line width of 3 mm. An open-stub, attached directly at the juncture where the vertical antenna element is attached to the transmission line, is used for fine tuning of the input impedance. The stub length has been adjusted manually (shortened) while measuring the reflection coefficient with a vector network analyzer (VNA). Open-stub matching is discussed in Chapter 5 in \cite{pozar_microwave_2011}. Furthermore the resonance frequency has been tuned by cutting the vertical antenna element piece by piece while measuring with the VNA until \ensuremath{S_{\text{11}}} had the minimum close to the desired frequency (Figure~\ref{fig:s11}\cp{Can we replace all these figures by PDFs?}). In the plot one can also see, that the antenna is quite narrowband\cp{We should report something like the 3dB bandwidth rather than saying something non-specific like this}. For backscattering this is not necessarily a disadvantage, since everything that is within the operating frequency of the antenna is backscattered. With our narrowband antenna the undesired backscattering of out-of-band signals is reduced. With our approach we were able to achieve an input impedance at 634 MHz of \ensuremath{Z_{in}=(52.2 + j 4.3) \Omega}. At 626 MHz \ensuremath{S_{\text{11}}} is still approximately -17 dB.

Besides the antenna, the transmitter consists of an \textit{MSP430} MCU, that sends data to the receiver. This transmitter controls an RF switch through an I/O pin. The switch can alternate the impedance connected to the antenna between matched (50 \ensuremath{\Omega}) and open. The software now steers this switch with a two 2 MHz rectangular timer signal to transmit a data 1 and is just leaves it open to transmit a data 0. Therefore the result is a frequency shift\cp{We need to explain this in the background} of the television signal by 2 MHz for a 1 and no shift for a 0. Hence a classical binary amplitude shift keying is the implemented modulation technique.

@@

\caption{Receiver architecture from a system point of view. Signal flow is from left to right.}

@@

C++ has been used to implement the receiver. We provide the code publicly available under \url{https://github.com/s3xm3x/backscatterBASKReceiver}. In figure \ref{fig:receiver_arch} we show the architecture of the receiver from an abstract point of view. A highly sophisticated bus system has been developed to exchange data between the different components of the system. The signal flow is from the left to the right. \textit{Librtlsdr} (cf. \cite{steve-m_librtlsdr}), which is based on \textit{libusb}, is used to transfer the data from the TV stick into our program. As described in section~\ref{sub:rtl2832}, the RTL2832 mixes down the high frequency to a intermediate frequency (both values are software adjustable). Subsequently the data is mixed with sine and cosine, low-pass filtered and finally transferred to the digital domain. This data is available as two 8-bit values. It is a well-known fact, that this technique, which is commonly known as I/Q demodulation, results in a decomposition of the signal into in-phase and quadrature components which are generally represented as real and imaginary parts respectively. With this two values, the phase and absolute value of the data can be determined for every sample. And can be seen as:

@@

	I+j\,Q = abs(I,Q) \cdot e^{j \times ang(I,Q)} 

@@

So the first block, which is entitled RTL2832U, provides the interface to the TV dongle. This block is represented as one object. It can set the frequency \ensuremath{f_{tuned}}, where the receiver is listening. Also it can set the sampling frequency \ensuremath{f_{samp}} and the analog gain of the tuner. The raw samples are pushed to the bus. 
The demodulator block is responsible for converting the sampled signals into ones and zeroes. Therefore the real and imaginary values first have to be deinterleaved out of the sample message. Now they are represented as integer values between 0 and 255 (8 bit). So one converts them to floats, whereas 127.5 equals 0. After that the values have to be downsampled. Usually we were sampling with 250 kHz. This is still much more than we need assuming an Additive White Gaussian Noise (AWGN) channel with a \ensuremath{S/N = 10\,dB} (cf. equation \ref{eq:awgn}). So we decided to reduce the sampling frequency to 25 kHz.

@@

A FIR filter naively implemented with floats has been used to achieve this. 10 kHz is the cut-off frequency of the first filter. All filters have been constructed with the internal filter design tool of GNU radio. After downsampling another filter is used to suppress noise. This filter has a cut-off frequency of 3 kHz. The last demodulation step is to rectify the signal (cf. equation \ref{eq:rectify}) and decide with a software-defined Schmitt trigger whether a sample is a 1 or a 0.

@@

These samples are broadcast on the bus. A registered listener receives the processed samples. This registered listener is the decoder, which is represented as an object as well. After the decoder received the samples it decides when a frame is sent or when the channel is idle. On the basis of a defined threshold the decoder determines whether a certain sample pattern of ones and zeros is a 1 bit or a 0 bit. The baudrate and the sampling rate lead to a certain bit duration. For example a sample rate \ensuremath{f_{sampling}} of 25 kHz combined with a baudrate of 1 kbit/s leads to 25 samples/bit. The decoder also includes a function to correlate the received bitstream with an expected pattern. Hence it is able to determine the bit error ratio (BER). 
Furthermore the code we have written so far subsumes a simulator to simulate ideal data coming out of the demodulator to test the algorithms behind the decoder. Also a simulator for the RTL2832U has been written to replay recorded samples, which facilitates debugging as well.    
